### 2019-03-28 paper/sex_selection_spacing-ver2.tex
Shorten first half of paper

Birth spacing has long served as a measure of son preference, with strong
son preference typically associated with shorter birth intervals the
fewer sons a family has \citep{ben-porath76b,Leung1988}.
However, the introduction of prenatal sex determination has the potential to
change the relationship between son preference and birth spacing.
The stronger the son preference, the more likely couples are to resort to sex-selective 
abortions, and each abortion automatically increases the interval between births by six 
months to a year.%
\footnote{
The increase consists of three parts.
First, starting from the time of the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Longer birth intervals in the absence of sons, which previously would be taken to indicate 
a lower son preference, now may arise precisely because a sufficiently strong son 
preference makes couples use sex selection.

In this paper, I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
I introduce and apply an empirical method that directly incorporates the effects of 
sex-selective abortions on the duration between births 
\emph{and} 
the likelihood of a son. 
The method can be used to analyze both situations with and without prenatal
sex selection.
I apply the method to birth histories of Hindu women, using data from the four
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2016.

% [Why India?]
India is a particularly compelling case since it has a long history of male preference, 
especially in the northern states \citep{Kishor1993,murthi95,arnold98}.
A significant proportion of Indian families exhibit differential stopping behavior, where, 
for a given number of children, they are more likely to have an additional child if they 
have not yet reached their preferred number of sons 
\citep{repetto72,Das1987,Arnold1997,arnold98,clark00,Basu2010,Barcellos2014}.%
\footnote{
Other countries show a similar pattern
\citep[see, for example,][]{larsen98,filmer09,Altindag2016}.
}
Moreover, mortality risk is higher for females than males for most ages groups, and
there has been an almost continuous increase in India's overall ratio of males to females 
over the last century \citep{dyson01,Navaneetham2011,Bongaarts2015}.

% [What do we know about use of sex selection?]
Over the last three decades, India has seen dramatic increases in the males-to-females 
ratio at birth as access to prenatal sex determination expanded 
\citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,jha06,Guilmoto2012,Portner2015b}.
Simultaneously, fertility has declined substantially, to the point where it is now at, or 
even below, replacement in some areas 
\citep{Guilmoto2013,Dharmalingam2014,International-Institute-for-Population-Sciences-IIPS2017}.
This decline in fertility, combined with a continued strong preference for sons, has 
intensified the pressure to use sex selection 
\citep{das_gupta97,Guilmoto2009a,Bongaarts2013,Jayachandran2017}.%
\footnote{
In some instances, however, the fertility decline appears to have created a stronger 
aversion to daughters rather than a stronger preference for son \citep{DiamondSmith2008}.
}
The strongest predictor of the use of sex selection is the sex composition of previous 
children; for families without a son, the higher the parity, the higher is the probability
of having a son as the next birth \citep{retherford03b,jha06,abrevaya09,Portner2015b}.
The use of sex selection also increases with socioeconomic status, especially 
education, and is more prevalent in cities than in rural areas, both possibly because 
of lower desired fertility \citep{retherford03b,jha06,Portner2015b}.

% [Why spacing?] 
There are four main reasons for examining birth spacing and how it changes with 
access to sex selection.
First, researchers have made extensive use of birth spacing as a measure of son preference, 
and it is critical to understand to what extent spacing is still useful as a measure of son 
preference.
In India, the sex of the last-born child has in the past significantly affected the 
duration to next birth, with the expected birth interval about three percent longer if the 
last-born was a son than if it was a daughter \citep{Bhalotra2008,Kumar2016}.
Ethnic Indians in South Africa also show a longer duration after the birth of a 
son than after a daughter \citep{Gangadharan2003}.%
\footnote{
Ethnic Indians in Malaysia do, however, show little evidence of son preference, although
the sample sizes are small \citep{Pong1994}.
}
In both Bangladesh and Vietnam, the more boys a family has the longer is the expected 
duration to the next birth \citep{Haughton1995,Haughton1996,Rahman1993,Soest2018}.
Outside Asia, the evidence is more mixed.
North Africa shows shorter spacing in the absence of sons, while a similar pattern does
not exist in Sub-Saharan Africa \citep{Rossi2015}.%
\footnote{
A possible exception is Senegal, where women with a higher risk of widowhood show
shorter birth spacing until they have secured a son \citep{Lambert2016}.
}

Second, birth spacing can have a substantial impact on children's outcomes.
Most of the research finds a negative impact of short spacing on child health---especially 
for very short intervals of 18 months or less---although identifying the causal mechanisms 
have proven more difficult \citep{Conde-Agudelo2006,Conde-Agudelo2012}.
There are three pathways through which birth intervals can affect health.
First, there is maternal depletion, where the mother does not have sufficient time to 
recover after her last pregnancy before she becomes pregnant again.%
\footnote{
This depletion can take the form of nutrition, folate, or cervical insufficiencies.
}
Second, disease transmission either from mother to child or between children.
Third, sibling competition, where siblings close in age compete for scarce resources and
parental care.
Which mechanism dominates is of particular interest when son preference drives spacing 
decisions.
Both the maternal depletion and the transmission of disease explanations predominately
affect the subsequent child and the mother.
Only sibling competition has the potential to affect the older child negatively.

In India, there is evidence that both the prior and the subsequent child have worse health 
outcomes and a higher likelihood of dying the shorter the interval between them, although 
the effects are not symmetrical
\citep{Whitworth2002,Bhargava2003,Rutstein2005,Maitra2008,Makepeace2008,Ghosh2018}.%
\footnote{
\cite{Bhalotra2008}, however, find that birth spacing only explains a limited amount
of neonatal mortality.
}
The increased mortality risk for prior children may come from shorter breastfeeding of 
girls to ensure that the mother can conceive again sooner, and with poor water
quality this shorter breastfeeding may result both in higher risk of dying and shorter
stature if the child survives \citep{Jayachandran2011,Jayachandran2017a}.
In other countries, focus has been mostly on the subsequent birth, but there is evidence 
in countries as diverse as Bangladesh, Brazil, El Salvadore, and Pakistan that
short intervals between births have adverse health effects
\citep{Cleland1984,Curtis1993,Davanzo2008,Gribble2009,Saha2013}.

The potential effects of spacing are not restricted to early outcomes, although
there is less evidence on longer-term effects and most comes from Western countries.
In the US, for example, a longer interval between births increases the older sibling's
test scores on the Peabody Individual Achievement Test, although there is no
statistically significant effect on the younger sibling \citep{Buckles2012}.
In line with this, close spacing appears to increase the risk of dropping out of
high school and lower the probability of attending a post-secondary school in both the
US and Sweden \citep{Powell1993,Pettersson-Lidbom2009}.
These results are, however, not uniformly supported, with recent results from Sweden 
showing no relationship between birth spacing, even if very short, and outcomes such as 
years of education completed, earnings, and unemployment \citep{Barclay2017}.

Third, the mother may also be affected by the length of the birth intervals.
The effect of spacing on mothers have received less attention than the effect on children, 
and reviews of the literature show mixed results on mother's anthropometric status 
\citep{Dewey2007,Conde-Agudelo2012}.
For India, there is, however, evidence that women with first-born girls are both
more likely to have short birth spacing and more likely to have anemia, possibly as
the result of the short spacing resulting in maternal depletion \citep{Milazzo2018}.
In Western countries, having a longer interval between births appear to have positive
effects on long-term labor market outcomes, such as participation and 
income \citep{Gough2017,Karimi2014}.

Finally, we know less about what determines spacing behavior in developing countries than 
in developed countries, and with declining fertility and increasing numbers of women 
entering the labor force in developing countries, understanding how couples make timing 
decisions will be necessary for the design of suitable policies \citep{Portner2018}.%
\footnote{
It is clear that birth spacing does respond to policies in both developed and 
developing countries \citep{Pettersson-Lidbom2009,Todd2012,Meckel2015,Ghosh2018}.
}
For example, the evidence on the effects of education and socio-economic status on
birth intervals is mixed, with some research showing either a positive association, no
association, or one that is changing over time 
\citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Kim2010,Soest2018}.

Even the effects of access to contraceptive and declining fertility on spacing are unclear.
On the one hand, access to contraceptives allows women to avoid too short spacing between 
birth, which would increase birth spacing.
On the other hand, increased reliability of access and effectiveness of contraceptives can 
lead to shorter intervals between births  \citep{Keyfitz1971,Heckman1976}.
With less reliable contraception parents choose a higher level of contraception, which 
results in longer spacing, to avoid having too many children by accident.
However, as contraception becomes more effective parents can more easily avoid future 
births, allowing them to reduce the spacing between births without having to worry about 
overshooting their preferred number of children.
These two counteracting effects may explain why better-educated women have shorter spacing 
than less educated women in some instances but not in others and why finding statistically 
significant effects of contraception use on birth spacing is difficult \citep{Yeakey2009}.%
\footnote{
There is evidence in Matlab that the effect of son preference on birth spacing is
stronger in areas with better access to family planning \citep{Rahman1993}.
Other evidence, however, points to families ability to time birth, even in the absence
of modern contraceptives \citep{Jayachandran2011,Alam2018}.
}

% [Hypotheses / questions]
Despite the increasing use of sex selection, the critical impacts of spacing on outcomes 
for both mother and children, and the extensive use of spacing to measure son 
preference in the literature, there has so far been no examination of what effect access 
to sex selection has had on spacing.
The overall hypothesis is that spacing patterns significantly changed as prenatal sex 
determination became available.
Specifically, I address three questions.

First, how did average birth intervals change over time for different groups and how are 
those changes related to the likely use of sex selection?
The underlying hypothesis is that in the absence of sex selection son preference leads to 
shorter spacing the fewer boys a family has for a given parity, whereas son preference 
increases the spacing the fewer boys a family has when prenatal sex selection is available.
I still expect parents to \emph{try} to conceive earlier after the birth of a girl than 
after a boy, but the abortions will increase the duration between births, and may even 
lead to longer average spacing after the birth of girls than after boys.
Families with son preference who---for one reason or another---do not use prenatal sex 
selection may continue to have shorter birth spacing.
Based on the prior literature on sex selection, I focus on how birth intervals have
changed for different education levels, urban or rural, and the sex composition of 
prior children.

Second, have the shapes of the survival curves changed?
Very short birth intervals are, as mentioned, associated with worse health outcomes and 
increased mortality risk \citep{Conde-Agudelo2012}.
Hence, understanding how the distribution of birth intervals changes can help us understand to 
what extent we can expect improvements in health outcomes.
We should, for example, expect different impacts on the distribution of birth intervals
depending on whether parents change their spacing behavior based on an increasing 
concern for child health or from a strong son preference leading to continued pregnancies
until they conceive a son.

Finally, how are the changes in birth intervals with increasing access to sex selection 
distributed geographically?
Given the size and cultural differences within India, it is unlikely that the increasing 
access to sex selection will have had the same effects throughout \citep{retherford03b}.
I, therefore, examine how birth intervals have changed across four regions.

Direct information on the use of sex selection is not available, but it is possible to 
compare different periods as a proxy since there have been substantial changes in access 
and legality of prenatal sex determination in India.
Abortion has been legal in India since 1971 and still is.
The first reports of sex determination appeared around 1982--83 
\citep{Sudha1999,bhat06,Grover2006}.
The number of clinics quickly increased, and knowledge about sex selection became 
widespread after a senior government official's wife aborted a fetus that turned out to be 
male \citep[p.\ 598]{Sudha1999}.
In 1994, the Central Government passed the Prenatal Diagnostic Techniques (PNDT) Act, 
making determining and communicating the sex of a fetus illegal.%
\footnote{
Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
The number of convictions has been low.
It took until January 2008 for the first state, Haryana, to reach five convictions.
Hence, private clinics operate with little risk of legal action \citep{Sudha1999}.
Furthermore, there is little evidence that bans like this significantly
affect sex ratios \citep{Das-Gupta2016}.
}

Based on these developments, I split the four decades into four periods.
The first period covers until prenatal sex determination became widely available, 
1972--1984.
The second period saw (mostly) unrestricted access to sex selection, 1985--1994.
Although, as reviewed above, there is evidence that there has been a substantial 
increase in the use of sex selection, some researchers argue that we might be close to or 
have already passed a turning-point in the use of sex selection 
\citep{Das_Gupta2009,Diamond-Smith2015}.
I, therefore, split the period from the implementation of the PNDT Act until the last 
survey into two, 1995--2004, and 2005--2016.

There are four main results.
First, there has been a general increase in the intervals between births for all education 
groups over the four decades covered by the data.
These increases are larger the higher the parity and the higher the education 
level.
Second, those women who are most likely to use sex selection, well-educated women with no
sons, have seen the largest increases in birth intervals.
As a result, we now have a complete reversal of the traditional spacing pattern for
well-educated women, and instead, observe the longest birth intervals for women with no 
sons.
Third, women who are least likely to use sex selection, those with no education in
rural areas, still follow the standard pattern of short spacing when they have girls and 
little evidence of the use of sex selection.
In other words, these women adhere to a strategy where they achieve a son through higher 
fertility rather than the use of sex selection.
Fourth, the traditional pattern of longer average spacing when women have one or more sons
appear to arise from more very long intervals, rather than from fewer very short birth 
intervals; across education groups there appear to be little difference in the likelihood
of having a very short birth interval across sex compositions, especially for lower
parities.

@@

The standard approach in the birth spacing literature is to use proportional hazard
models with a single exit, the birth of a child.
There are two problems with the standard approach in this setting.
First, and most importantly, the introduction of sex selection means that the sex of the 
next child is no longer necessarily random and that parents' choices will impact the 
spacing to the birth of a girl or a boy differently.
I, therefore, use a competing risk set-up, which can capture both the non-randomness of
the birth outcome and the differential spacing to the birth of a girl or a boy.
To my knowledge, this is the first application of competing risk models to birth
spacing.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

Second, it is unlikely---even in the absence of prenatal sex determination---that 
characteristics, such as the sex composition of previous births, have the same effects 
throughout the entire spell as is assumed by the proportional hazard model.
Assuming that the effect of sex composition is the same throughout a spell is especially 
problematic for higher-order spells where different sex composition of previous births can 
lead to substantial differences both in the likelihood of progressing to the next birth 
and how soon couples want their next child if they are going to have one.
The introduction of prenatal sex determination exacerbates any bias from the 
proportionality assumption since sex-selective abortion use varies across groups,
which affects birth spacing, and because a household's use of sex selection may vary 
within a spell, which means that the effects of covariates vary within the spell as well.
I, therefore, use a non-proportional hazard specification, which allows the shape of the 
hazard functions to vary across groups.
The use of a non-proportional specification also mitigates any potential effects 
of unobserved heterogeneity when used in conjunction with a flexible baseline hazard 
\citep{Dolton1995}.

@@

Equation (\ref{eq:hazard}) is equivalent to the logistic hazard model and has the same 
likelihood function as the multinomial logit model \citep{allison82,jenkins95}.
Hence, transforming the data, so each observation is an interval---here equal
to three months---the model can be estimated using a standard multinomial logit model.%
\footnote{
The multinomial model assumes that alternative exit states are stochastically independent
(Independence of Irrelevant Alternatives or IIA), which rules out any individual-specific 
unmeasured or unobservable factors that affect both the hazard of having a girl and the 
hazard of having a boy.
I, therefore, include a proxy for fecundity discussed in Section \ref{sec:data}.
Also, the multivariate probit model can be used as an alternative because the IIA is not 
imposed \citep{han90}.
The results are substantially identical between these two models and
available upon request.
}
In the reorganized data the outcome variable is 0 if the woman does not have a child in a 
given interval (the base outcome), 1 if she gives birth to a son in that interval, and 2 
if she gives birth to a daughter in that interval.

@@

The data come from the four rounds of the National Family Health Survey 
(NFHS-1, NFHS-2, NFHS-3, and NFHS-4),
collected in 1992--1993, 1998--1999, 2005--2006, and 2015--2016.%
\footnote{
A delay in the survey for Tripura means that NFHS-2 has some observation 
collected in 2000.
}
The surveys are large: NFHS-1 covered 89,777 ever-married women 
aged 13--49 from 88,562 households;
NFHS-2 covered 90,303 ever-married women aged 15--49 from 92,486 households;
NFHS-3 covered 124,385 never-married and ever-married women aged 
15--49 from 109,041 households;
and 
NFHS-4 covered 699,686 never-married and ever-married women aged
15--49 from 601,509 households surveyed.

I exclude visitors to the household, as well as
women who have been married more than once, divorced, or who are 
not living with their husband,
women with inconsistent age at marriage,
and those with missing information on education.
Women interviewed in NFHS-3 or NFHS-4 who were never married or where Gauna 
had not yet been performed were also dropped.
The same goes for women who had at least one multiple births,
reported having a birth before age 12, had a birth before marriage, or an interval between 
births of less than nine months.

Finally, I restrict the sample to Hindus,
who constitute about 80\% of India's population.
If the use of sex selection differ between Hindu and other religions, such 
as Sikhs, assuming that the baseline hazard is the same would lead to bias.
The other groups are each so small relative to Hindus that it is not
possible to estimate different baseline hazards for each group.
Furthermore, the other groups are so different in background and son 
preference that combining them into one group would not make sense.

There are four advantages to using the NFHS.
First, surveys enumerators pay careful attention to spacing between births and
probe for ``missed'' births.
Second, no other surveys cover as extended a period in the same amount of detail.
The four NFHS rounds allow me to show how spacing and 
sex ratio changed from before sex-selective abortions were available until 2016.
Third, NFHS has birth histories for a large number of women.
% %
% \footnote{
% The Special Fertility and Mortality Survey appears to cover a much large number of households
% than the three rounds of the NFHS combined, but \citet{jha06} only use the births that 
% took place in 1997 making their sample sizes by parity smaller than here.
% Their sample consists of 133,738 births of which 38,177 were first-born, 
% 36,241 second-born, and 23,756 were third-born.
% }
Finally, even if probing for missing births does not eliminate recall error,   
the overlap in cohorts covered and the large sample size makes it possible 
to establish where recall error remains a problem.

Recall error arises mainly from child mortality when respondents are 
reluctant to discuss deceased children.
Systematic recall error, where the likelihood of reporting a deceased 
child depends on the sex of the child, is especially problematic because 
it biases both spacing and sex ratios.
Probing catches many missed births, but systematic recall error is 
still a potentially substantial problem.

Three factors contribute to the problem here.
First, girls have a significantly higher mortality risk than boys.
Second, son preference may increase the probability that parents recall boys 
more readily than girls.
Finally, in NFHS-1 and NFHS-2 enumerators probed only for a missed birth if the
initially reported birth interval was four calendar years or more,
but, given short durations between births, especially after the birth of a girl,
that procedure is unlikely to pick up all missed children.

Recall error is heavily dependent on how long ago a woman was married.%
\footnote{
See the discussion in the online Appendix.
}
I, therefore, drop women married 22 years or more.
The final sample consists of 
\input{../tables/num_women.tex}
% [number] women, with [number] parity one through four births.

\subsection{Spell Definition\label{sec:spell_def}}

I focus here on the second through fourth spells, that is, on the intervals from the first 
birth until at most the fourth birth.%
\footnote{
I exclude the first spell from marriage to the first spell because of data issues.
In NFHS-1 and NFHS-2 the only information on marriage timing is the age at marriage, and 
all spell durations are therefore imputed.
In NFHS-3 about a third of respondent did not provide complete information on the
timing of marriage and therefore had the length of their first spell imputed using only
the age at marriage.
The result of the imputation is that large numbers of women end up with very short,
i.e., less than nine months, or negative spells and that comparison of spell duration over 
time is subject to an excessive amount of noise.
}
The spells all begin nine months after the previous birth 
because that is the earliest we should expect to observe a new birth, which also means 
that they capture the interpregnancy interval for pregnancies carried to term.
A spell continues until either a child is born or censoring occurs.
Censoring can happen for three reasons:
the survey takes place;
sterilization of the woman or her husband;
or too few births are observed for the method to work.
For all spells censoring is set at 96 months (eight years) after a birth can occur.
With these cut-offs, less than one percent of observed births occur after
the spell cut-off.%
\footnote{
\input{../tables/num_missed.tex}
}

I group spells into four periods based on the year of the previous birth:
1972--1984, 1985--1994, 1995--2004, and 2005--2016.
The periods follow the main changes in the availability and legality of 
prenatal sex determination discussed in the Introduction.
The allocation into periods is determined by when conception and therefore 
decisions on sex selection can begin, even if we do not observe
any births until nine months later.

>>>>

Birth spacing has long served as a measure of son preference, with strong
son preference associated with shorter birth intervals the
fewer sons a family has \citep{ben-porath76b,Leung1988}.
However, the introduction of prenatal sex determination potentially 
changed the relationship between son preference and birth spacing.
The stronger the son preference, the more likely couples are to resort to sex-selective 
abortions, and each abortion increases the interval between births by six to twelve 
months.%
\footnote{
The increase consists of three parts.
First, starting from the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Hence, longer birth intervals in the absence of sons, which previously would be taken to 
indicate a lower son preference, now may instead arise because couples with a strong son 
preference use sex selection.

Here, I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
I introduce an empirical method that directly incorporates the effects of 
sex-selective abortions on the duration between births 
\emph{and} 
the likelihood of a son. 
The method can be used to analyze both situations with and without prenatal
sex selection.
I apply the method to birth histories of Hindu women, using data from the four
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2016.

% [Why India?]
India is a particularly compelling case with its long history of male preference, 
especially in the northern states \citep{Kishor1993,murthi95,arnold98}.
Mortality risk is higher for females than males for most ages groups, resulting in
an almost continuous increase in India's overall ratio of males to females 
over the last century \citep{dyson01,Navaneetham2011,Bongaarts2015}.
Differential stopping behavior is widespread, with an additional child more likely the
fewer sons a family has 
\citep{repetto72,Das1987,Arnold1997,arnold98,clark00,Basu2010,Barcellos2014}.%
\footnote{
Other countries show a similar pattern
\citep[see, for example,][]{larsen98,filmer09,Altindag2016}.
}
Furthermore, fertility has declined substantially, to the point where it is now at, or even 
below, replacement in some areas 
\citep{Guilmoto2013,Dharmalingam2014,International-Institute-for-Population-Sciences-IIPS2017}.
The combination of falling fertility, strong son preference, and expanding access to 
prenatal sex determination has lead to dramatic increases in the males-to-females ratio at 
birth in India from sex selection over the last three decades
\citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,jha06,Guilmoto2009a,Guilmoto2012,Bongaarts2013,Portner2015b,Jayachandran2017}.

% [Why spacing?] 
There are four main reasons for examining birth spacing and how it changes with 
access to sex selection.
First, researchers have made extensive use of birth spacing as a measure of son preference, 
and it is critical to understand to what extent spacing is still a useful measure of son 
preference.
For example, in Bangladesh, India, and Vietnam, having more boys, and particularly having 
a boy as the last-born child, significantly increase the duration to next birth,
\citep{Haughton1995,Haughton1996,Rahman1993,Bhalotra2008,Kumar2016,Soest2018}.%
\footnote{
Ethnic Indians in South Africa also show a longer duration after the birth of a 
son than after a daughter \citep{Gangadharan2003}.
}
Outside Asia, the evidence is more mixed, with North Africa showing shorter spacing in the 
absence of sons, while a similar pattern does not exist in Sub-Saharan Africa 
\citep{Rossi2015}.

Second, birth spacing can affect children's short and long-term outcomes.
Most of the research finds a negative impact of short spacing on child health, especially 
for very short intervals of 18 months or less \citep{Conde-Agudelo2006,Conde-Agudelo2012}.
There is evidence in countries as diverse as Bangladesh, Brazil, El Salvadore, India,
and Pakistan of worse health outcomes and a higher mortality risk the shorter the birth
interval
\citep{Cleland1984,Curtis1993,Whitworth2002,Bhargava2003,Rutstein2005,Bhalotra2008,Davanzo2008,Maitra2008,Makepeace2008,Gribble2009,Jayachandran2011,Saha2013,Jayachandran2017a,Ghosh2018}.
Most of the evidence on longer-term outcomes come from Western countries.
In the US, for example, a longer interval between births increases the older, but not the
younger, sibling's test scores \citep{Buckles2012}.
Furthermore, close spacing appears to increase the risk of dropping out of
high school and lower the probability of attending a post-secondary school in both the
US and Sweden \citep{Powell1993,Pettersson-Lidbom2009}.%
\footnote{
However, recent results from Sweden show no relationship between birth spacing, even if 
very short, and outcomes such as years of education completed, earnings, and unemployment 
\citep{Barclay2017}.
}

Third, the mother may also be affected by the length of the birth intervals, although 
reviews of the literature show mixed results on mother's anthropometric status 
\citep{Dewey2007,Conde-Agudelo2012}.
For India, there is, however, evidence that women with first-born girls are more likely to 
have anemia, possibly because short birth spacing can resulting in maternal depletion 
\citep{Milazzo2018}.
In Western countries, a longer interval between births has positive effects on long-term 
labor market outcomes, such as participation and income \citep{Gough2017,Karimi2014}.

Finally, we know less about what determines spacing behavior in developing countries than 
in developed countries, and with declining fertility and increasing numbers of women 
entering the labor force, understanding how couples make timing decisions is necessary 
for the design of suitable policies \citep{Portner2018}.%
\footnote{
It is clear that birth spacing does respond to policies in both developed and 
developing countries \citep{Pettersson-Lidbom2009,Todd2012,Meckel2015,Ghosh2018}.
}
Even the effects of access to contraceptive and declining fertility on spacing are unclear.
On the one hand, access to contraceptives allows women to avoid too short spacing between 
birth.
On the other hand, increased reliability of access and effectiveness of contraceptives can 
lead to shorter intervals between births if women used to have longer spacing to avoid
having too many children by accident \citep{Keyfitz1971,Heckman1976}.
These two counteracting effects may explain why better-educated women have shorter spacing 
than less educated women in some instances but not in others and why finding statistically 
significant effects of contraception use on birth spacing is difficult 
\citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Yeakey2009,Kim2010,Soest2018}.%
\footnote{
In Matlab, the effect of son preference on birth spacing is stronger in areas with better 
access to family planning \citep{Rahman1993}.
Other evidence, however, points to families ability to time births, even in the absence
of modern contraceptives \citep{Jayachandran2011,Alam2018}.
}

% [Hypotheses / questions]
Despite the increasing use of sex selection, the critical impacts of spacing on outcomes 
for both mother and children, and the extensive use of spacing to measure son preference, 
there has so far been no systematic examination of what effect access to sex selection 
have on spacing.
I address three questions to remedy this shortcoming.

First, how did average birth intervals change over time for different groups and how do 
these changes relate to the use of sex selection?
The hypothesis is that in the absence of sex selection son preference leads to shorter 
spacing the fewer boys a family has, whereas son preference increases spacing the fewer 
boys a family has when prenatal sex selection is available, possibly to the point where
we observe a reversal of the standard spacing pattern.
Families with son preference who---for one reason or another---do not use prenatal sex 
selection likely continue to have shorter birth spacing.
Based on the prior literature on sex selection, I focus on how birth intervals have
changed for different education levels, urban or rural, and the sex composition of 
prior children.

Second, have the shapes of the survival curves changed?
Although examining average birth intervals helps us understand the overall changes in 
spacing behavior, it tells us little about how these changes came about.
For example, the motivation behind an increase in average spacing is different if it
comes mainly from parents trying to avoid very short birth intervals, which, as mentioned,
are associated with worse child outcomes, or from continued use of sex selection to
ensure a son.

Finally, how are the changes in birth intervals distributed geographically?
Given the cultural differences within India, it is unlikely that the increasing 
access to sex selection will have had the same effects throughout \citep{retherford03b}.
I, therefore, examine how birth intervals have changed across four regions.

There are four main results.
First, the intervals between births have increased for all education groups over the four 
decades.
These increases are larger the higher the parity and the higher the education 
level.
Second, those women who are most likely to use sex selection---well-educated women with no
sons---have seen the largest increases in birth intervals.
As a result, well-educated women now show a complete reversal of the traditional spacing 
pattern, with the longest birth intervals for women with no sons.
Third, women who are least likely to use sex selection, those with no education in
rural areas, still follow the standard pattern of short spacing when they have girls and 
little evidence of the use of sex selection.
Fourth, the traditional pattern of longer average spacing with sons appear to arise from 
more very long intervals, rather than from fewer very short birth intervals; across 
education groups, there appears to be little difference in the likelihood of having a very 
short birth interval across sex compositions, especially for lower parities.

@@

The standard approach in the birth spacing literature is to use proportional hazard
models with a single exit, the birth of a child.
There are two problems with this approach in this setting.
First, and most importantly, the introduction of sex selection means that the sex of the 
next child is no longer necessarily random and that parents' choices will impact the 
spacing to the birth of a girl or a boy differently.
I, therefore, use a competing risk set-up, which can capture both the non-randomness of
the birth outcome and the differential spacing.
To my knowledge, this is the first application of competing risk models to birth
spacing.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

Second, it is unlikely---even in the absence of prenatal sex determination---that 
characteristics, such as the sex composition of previous births, have the same effects 
throughout the entire spell as is assumed by the proportional hazard model.
The proportionality assumption is especially problematic for higher-order spells where 
there are substantial differences across groups both in the likelihood of progressing to 
the next birth and how soon couples want their next child if they are going to have one.
The introduction of prenatal sex determination exacerbates any bias from the 
proportionality assumption since sex-selective abortion use varies across groups,
which affects birth spacing, and because a household's use of sex selection may vary 
within a spell, which means that the effects of covariates vary within the spell as well.
I, therefore, use a non-proportional hazard specification, which allows the shape of the 
hazard functions to vary across groups.
The use of a non-proportional specification also mitigates any potential effects 
of unobserved heterogeneity when used in conjunction with a flexible baseline hazard 
\citep{Dolton1995}.

@@

Equation (\ref{eq:hazard}) is equivalent to the logistic hazard model and has the same 
likelihood function as the multinomial logit model \citep{allison82,jenkins95}.
Hence, transforming the data, so each observation is an interval---here equal
to three months---the model can be estimated using a standard multinomial logit model.
%
% \footnote{
% The multinomial model assumes that alternative exit states are stochastically independent
% (Independence of Irrelevant Alternatives or IIA), which rules out any individual-specific 
% unmeasured or unobservable factors that affect both the hazard of having a girl and the 
% hazard of having a boy.
% I, therefore, include a proxy for fecundity discussed in Section \ref{sec:data}.
% Also, the multivariate probit model can be used as an alternative because the IIA is not 
% imposed \citep{han90}.
% The results are substantially identical between these two models and
% available upon request.
% }
In the reorganized data the outcome variable is 0 if the woman does not have a child in a 
given interval (the base outcome), 1 if she gives birth to a son in that interval, and 2 
if she gives birth to a daughter in that interval.

@@

The data come from the four rounds of the National Family Health Survey
collected in 1992--1993, 1998--1999, 2005--2006, and 2015--2016.%
\footnote{
A delay in the survey for Tripura means that NFHS-2 has some observation 
collected in 2000.
}
The surveys are large: NFHS-1 covered 89,777 ever-married women 
aged 13--49 from 88,562 households;
NFHS-2 covered 90,303 ever-married women aged 15--49 from 92,486 households;
NFHS-3 covered 124,385 never-married and ever-married women aged 
15--49 from 109,041 households;
and 
NFHS-4 covered 699,686 never-married and ever-married women aged
15--49 from 601,509 households surveyed.

I exclude visitors to the household, as well as women who fall into any of the following 
categories: never married; no gauna yet; married more than once; divorced; 
not living with husband; inconsistent age at marriage; or education information missing.  
The same goes for women who had at least one multiple births, reported having a birth 
before age 12, had a birth before marriage, or an interval between births of less than 
nine months. 
Finally, I restrict the sample to Hindus, who constitute about 80\% of India’s population.%
\footnote{
If the use of sex selection differ between Hindu and other groups assuming
that the baseline hazard is the same would lead to bias. The other
groups are each so small that it is not possible to estimate different
baseline hazards for each group, and are so different that combining
them into one group would not make sense.
}

In addition to the large number of women surveyed and the long period covered, one of the
benefits of the NFHS is that surveys enumerators pay careful attention to spacing between 
births and probe for ``missed'' births.
Nevertheless, systematic recall error, where the likelihood of reporting a deceased 
child depends on the sex of the child, is still a potentially substantial problem and
will bias both spacing and sex ratios.
As I show in the online Appendix, recall error is heavily dependent on how long ago a 
woman was married, and I, therefore, drop women married 22 years or more.
The final sample consists of 
\input{../tables/num_women.tex}
% [number] women, with [number] parity one through four births.

Direct information on the use of sex selection is not available, so I compare different 
periods as a proxy, based on the changes in access and legality of prenatal sex 
determination in India.
Abortion has been legal in India since 1971.
Reports of sex determination appeared around 1982--83, and the number of clinics 
quickly increased \citep{Sudha1999,bhat06,Grover2006}.
In 1994, the Prenatal Diagnostic Techniques (PNDT) Act made determining and communicating 
the sex of a fetus illegal.%
\footnote{
Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
There is little evidence that the ban significantly affected sex ratios \citep{Das-Gupta2016}.
}
Although the use of sex selection increased even after 1994, some researchers argue that 
we have passed a turning point in its use \citep{Das_Gupta2009,Diamond-Smith2015}.
I, therefore, use four periods: 1972--1984, 1985--1994, 1995--2004, and 2005--2016.
The allocation of spells into periods is determined by when conception, and therefore 
decisions on sex selection, can begin, even if we do not observe any births until nine 
months later. 
This allocation mechanism means that some spells cover two periods, which may bias 
downward the differences between the periods.

I focus here on the second through fourth spells, that is, on the intervals from the first 
birth until at most the fourth birth.
I exclude the spell from marriage to the first birth because of data issues.
In NFHS-1 and NFHS-2 the only information on marriage is the age at marriage, and 
all first spell durations are therefore imputed.
In NFHS-3 about a third of respondent did not provide complete information on the
timing of marriage and had the length of their first spell imputed using the age at 
marriage.
The imputation results in large numbers of women with spells that are either negative or 
less than nine months and makes a comparison of spell duration over time subject to an 
excessive amount of noise.

The spells all begin nine months after the previous birth, which is the earliest we should 
expect to observe a new birth.
A spell continues until either a child is born or censoring occurs.
Censoring can happen for three reasons:
the survey takes place;
sterilization of the woman or her husband;
or too few births are observed for the method to work.
For all spells censoring is set at 96 months (eight years) after a birth can occur.
With these cut-offs, less than one percent of observed births occur after
the spell cut-off.%
\footnote{
\input{../tables/num_missed.tex}
}
